---
title: "Test-Great Tit"
author: "Wenyi Liu"
date: "2025-10-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the packages required and set seed
```{r}
require(doBy)
require(car)
require(mgcv) 

rm(list=ls(all=TRUE))

set.seed(123)
```

# Process the dataset

```{r}
# load the dataset
great_tit <- read.csv("Great Tit.csv")

# check the head of the columns
head(great_tit)

# expected number of chicks/female: ex = noffspring / Nx
great_tit$ex <- (great_tit$noffspring / great_tit$Nx)*2

# Survival: sx =N(x+1)/Nx The probability of indivduals with age x survive to x+1
great_tit$sx <- c(great_tit$Nx[-1] / great_tit$Nx[-nrow(great_tit)], NA)

# change NA values to zero
ex_vec <- ifelse(is.na(great_tit$ex), 0, great_tit$ex)
sx_vec <- ifelse(is.na(great_tit$sx), 0, great_tit$sx)

#Check the data
print(great_tit)

```

```{r}
#This is another dataset contains information about chick size and chick death rate
breeding_data <- read.csv("breeding_data.csv")

#Filter out great tits from the whole breeding data
breeding_gt <- subset(breeding_data, Species == "g")

# check the head of the table
head(breeding_gt)

#remove rows contains NA in num. chicks
library(dplyr)
breeding_gt <- breeding_gt %>%
  filter(!is.na(Num.chicks))

#check whether there is still NA in the filtered dataset
sum(is.na(breeding_gt$Num.chicks))  # should be 0

#calculate mean no. of chicks (succeed clutch size?)
mean_chick <- mean(breeding_gt$Num.fledglings, na.rm = TRUE)
mean_chick

#fecundity=ex/mean_chick
fx_vec <- ex_vec/mean_chick

```

# Set working directory
```{r}
setwd("D:/Nian/Oxford/UG/Year 4/Year 4 project/Method/MethodTest_GreatTit")
## run the utility functions
source("./Standard Graphical Pars.R")

## run the ungulate IBM
source("./Ungulate Demog Funs_GT.R")
```

#Set the simulation parameters and run the simulation
```{r}
## set the simulation parameters
init.pop.size <- 5000   
n.yrs <- 10

## run the ungulate IBM
source("./Ungulate Simulate IBM_GT.R")

## print the final population size
cat("Final population size:", tail(pop.size.t, 1), "\n")

```
take a random sample of 100 observations
we need to take a random sample because new recruits are placed at the start of the data frame 
```{r}
rows.to.keep <- sample.int(nrow(sim.data), 5000)
sim.data <- sim.data[rows.to.keep,]
```

##Section 2: Plot density~lifespan and density~LRO

#Density~lifespan

```{r}
# install.packages(c("dplyr", "ggplot2", "survival", "flexsurv"))

library(dplyr)
library(ggplot2)
library(survival)
library(flexsurv)

# 1) Build survival data (include right-censored individuals)
dat_surv <- sim.data %>%
  arrange(ID, Year, Age) %>%
  group_by(ID) %>%
  summarise(
    # first observed age
    first_age = min(Age, na.rm = TRUE),
    # the age when Surv == 0 appears for the first time (death event)
    death_age_event = if (any(Surv == 0, na.rm = TRUE)) Age[which(Surv == 0)[1]] else NA_real_,
    # last observed age and Surv status
    last_age = last(Age),
    last_surv = last(Surv),
    .groups = "drop"
  ) %>%
  mutate(
    # event time: if death observed, (death_age + 1 - first_age)
    # otherwise right-censored at (last_age + 1 if still survived)
    time = ifelse(
      !is.na(death_age_event),
      (death_age_event + 1) - first_age,
      (last_age + ifelse(last_surv == 1, 1, 0)) - first_age
    ),
    # status = 1 for observed death, 0 for right-censored
    status = ifelse(!is.na(death_age_event), 1, 0)
  ) %>%
  filter(is.finite(time), time > 0)

# 2) Fit several parametric survival distributions
fits <- list(
  weibull = flexsurvreg(Surv(time, status) ~ 1, data = dat_surv, dist = "weibull"),
  lnorm   = flexsurvreg(Surv(time, status) ~ 1, data = dat_surv, dist = "lognormal"),
  gamma   = flexsurvreg(Surv(time, status) ~ 1, data = dat_surv, dist = "gamma")
)

# 3) Select the best model by AIC
aics <- sapply(fits, AIC)
best_name <- names(aics)[which.min(aics)]
best_fit  <- fits[[best_name]]
cat("Best distribution by AIC:", best_name, "\n")

# 4) Create a fine grid of time values for plotting
t_grid <- seq(min(dat_surv$time), max(dat_surv$time), length.out = 400)

# 5) Compute the density (smooth lifespan distribution curve)
get_density <- function(dist_name, fit, t) {
  pars <- as.list(fit$res[,"est"])
  switch(dist_name,
    weibull = dweibull(t, shape = pars$shape, scale = pars$scale),
    lnorm   = dlnorm(t, meanlog = pars$meanlog, sdlog = pars$sdlog),
    gamma   = dgamma(t, shape = pars$shape, rate = pars$rate),
    stop("Unsupported distribution")
  )
}

dens_vals <- get_density(best_name, best_fit, t_grid)
df_plot <- data.frame(time = t_grid, density = dens_vals)

# 6) (Optional) Kaplan-Meier as a reference (step curve)
km <- survfit(Surv(time, status) ~ 1, data = dat_surv)
km_df <- data.frame(
  time = km$time,
  surv = km$surv
)
km_df$cdf <- 1 - km_df$surv  # convert to cumulative distribution

# 7) Plot smooth lifespan density curve
p_density <- ggplot(df_plot, aes(x = time, y = density)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Lifespan (years)", 
    y = "Density", 
    title = paste0("Smoothed lifespan distribution (", best_name, ", AIC=", round(min(aics),1), ")")
  ) +
  theme_minimal()

p_density

```

#Density~LRO
```{r}
####Remain to be done####

```





##Section 3: Model diagnostics for age-based survival and fecundity functions##

load the packages
```{r}
library(mgcv)
library(ggplot2)
```

#SURVIVAL DIAGNOSTICS
```{r}
cat("=== SURVIVAL DIAGNOSTICS ===\n")

# Fit a logistic regression model (GLM) and a smooth model (GAM)
mod.surv <- glm(Surv ~ Age, family = binomial, data = sim.data)
gam.surv <- gam(Surv ~ s(Age, k = length(unique(sim.data$Age)) - 1),
                family = binomial, data = sim.data, method = "REML")

summary(mod.surv)
summary(gam.surv)

# Predict fitted values for plotting
ages_seq <- seq(min(sim.data$Age), max(sim.data$Age))
pred_glm <- predict(mod.surv, newdata = data.frame(Age = ages_seq), type = "response")
pred_gam <- predict(gam.surv, newdata = data.frame(Age = ages_seq), type = "response")

# Plot observed vs fitted survival
plot(sim.data$Age, sim.data$Surv, pch = 1, cex = 0.7,
     xlab = "Age", ylab = "Survival probability",
     main = "Survival ~ Age diagnostics", ylim = c(0,1))
lines(ages_seq, pred_glm, col = "blue", lwd = 2)
lines(ages_seq, pred_gam, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("GLM (logistic)", "GAM (smooth)"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))

#Plot fitted VS observed residuals
plot(fitted(mod.surv), residuals(mod.surv, type = "deviance"),
     xlab = "Fitted values", ylab = "Deviance residuals")
abline(h = 0, col = "red", lty = 2)


# Basic diagnostics and model comparison
AIC(mod.surv, gam.surv)
anova(mod.surv, gam.surv, test = "Chisq")


```

#FECUNDITY DIAGNOSTICS
```{r}
cat("=== FECUNDITY DIAGNOSTICS ===\n")

# Fit Poisson regression models for fecundity (count data)
mod.fec <- glm(Offspring ~ Age, family = poisson, data = sim.data)
gam.fec <- gam(Offspring ~ s(Age, k = length(unique(sim.data$Age)) - 1),
               family = poisson, data = sim.data, method = "REML")

summary(mod.fec)
summary(gam.fec)

# Predict fitted values for plotting
pred_glm_f <- predict(mod.fec, newdata = data.frame(Age = ages_seq), type = "response")
pred_gam_f <- predict(gam.fec, newdata = data.frame(Age = ages_seq), type = "response")

# Plot observed vs fitted fecundity
plot(sim.data$Age, sim.data$Offspring, pch = 1, cex = 0.7,
     xlab = "Age", ylab = "Fecundity",
     main = "Fecundity ~ Age diagnostics")
lines(ages_seq, pred_glm_f, col = "blue", lwd = 2)
lines(ages_seq, pred_gam_f, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("GLM (Poisson)", "GAM (smooth)"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))

#Plot fitted VS observed residuals
plot(fitted(mod.surv), residuals(mod.surv, type = "deviance"),
     xlab = "Fitted values", ylab = "Deviance residuals")
abline(h = 0, col = "red", lty = 2)

# Model comparison
shapiro.test(residuals(mod.surv, type = "deviance"))
AIC(mod.fec, gam.fec)
anova(mod.fec, gam.fec, test = "Chisq")
```

