---
title: "Test-Great Tit"
author: "Wenyi Liu"
date: "2025-10-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the packages required and set seed
```{r}
require(doBy)
require(car)
require(mgcv) 

rm(list=ls(all=TRUE))

set.seed(123)
```

# Process the dataset

```{r}
# load the dataset
great_tit <- read.csv("Great Tit.csv")

# check the head of the columns
head(great_tit)

# Survival: sx =N(x+1)/Nx The probability of indivduals with age x survive to x+1
great_tit$sx <- c(great_tit$Nx[-1] / great_tit$Nx[-nrow(great_tit)], NA)

# expected number of chicks/female: ex = noffspring / Nx
great_tit$fx <- (great_tit$noffspring / great_tit$Nx)*2/great_tit$sx



# change NA values to zero
fx_vec <- ifelse(is.na(great_tit$fx), 0, great_tit$fx)
sx_vec <- ifelse(is.na(great_tit$sx), 0, great_tit$sx)

#Check the data
print(great_tit)

```


## Set working directory

```{r}
setwd("D:/Nian/Oxford/UG/Year 4/Year 4 project/Method/MethodTest_GreatTit")
## run the utility functions
source("./Standard Graphical Pars.R")

## run the ungulate IBM
source("./Demog Funs_GT.R")
```

# Set the simulation parameters and run the simulation

```{r}

## run the ungulate IBM
source("./Simulate IBM_GT.R")
sim.data <- simulate_ibm(sx_vec = sx_vec,
                    fx_vec = fx_vec,
                    great_tit = great_tit,
                    init.pop.size = 5000,
                    n.yrs = 10,
                    prop_female = 0.5,
                    max.pop.size = 8000,
                    return_history = TRUE,
                    seed = 123
                    )

```

## Run the counterfactual model
```{r}
source("./Plateau fun.R")

## Fot great tit, sexual maturity occurs at age = 1 
maturity_age <- 1L

## Create plateaued versions of fx_vec and sx_vec
fx_cf <- plateau_after_maturity(fx_vec, maturity_age, is_survival = FALSE)
sx_cf <- plateau_after_maturity(sx_vec, maturity_age, is_survival = TRUE)

## Feed them into your main simulation
cf.data <- simulate_ibm(sx_vec = sx_cf, fx_vec = fx_cf, 
                        great_tit = great_tit, init.pop.size = 5000, n.yrs = 10)

```

take a random sample of 5000 observations
we need to take a random sample because new recruits are placed at the start of the data frame 
```{r}
rows.to.keep <- sample.int(nrow(sim.data), 5000)
sim.data <- sim.data[rows.to.keep,]
cf.data <-cf.data[rows.to.keep,]
```



# Section 2: Plot density~lifespan and density~LRO

#Density~lifespan
  #Simulation data

```{r}
# install.packages(c("dplyr", "ggplot2", "survival", "flexsurv"))

library(dplyr)
library(ggplot2)
library(survival)
library(flexsurv)

# 1) Build survival data (include right-censored individuals)
sim.surv <- sim.data %>%
  arrange(ID, Year, Age) %>%
  group_by(ID) %>%
  summarise(
    # first observed age
    first_age = min(Age, na.rm = TRUE),
    # the age when Surv == 0 appears for the first time (death event)
    death_age_event = if (any(Surv == 0, na.rm = TRUE)) Age[which(Surv == 0)[1]] else NA_real_,
    # last observed age and Surv status
    last_age = last(Age),
    last_surv = last(Surv),
    .groups = "drop"
  ) %>%
  mutate(
    # event time: if death observed, (death_age + 1 - first_age)
    # otherwise right-censored at (last_age + 1 if still survived)
    time = ifelse(
      !is.na(death_age_event),
      (death_age_event + 1) - first_age,
      (last_age + ifelse(last_surv == 1, 1, 0)) - first_age
    ),
    # status = 1 for observed death, 0 for right-censored
    status = ifelse(!is.na(death_age_event), 1, 0)
  ) %>%
  filter(is.finite(time), time > 0)

# 2) Fit several parametric survival distributions
sim.fits <- list(
  weibull = flexsurvreg(Surv(time, status) ~ 1, data = sim.surv, dist = "weibull"),
  lnorm   = flexsurvreg(Surv(time, status) ~ 1, data = sim.surv, dist = "lognormal"),
  gamma   = flexsurvreg(Surv(time, status) ~ 1, data = sim.surv, dist = "gamma")
)

# 3) Select the best model by AIC
sim.aics <- sapply(sim.fits, AIC)
sim.best.name <- names(sim.aics)[which.min(sim.aics)]
sim.best.fit  <- sim.fits[[sim.best.name]]
cat("Best distribution by AIC:", sim.best.name, "\n")

# 4) Create a fine grid of time values for plotting
sim.t.grid <- seq(min(sim.surv$time), max(sim.surv$time), length.out = 400)

# 5) Compute the density (smooth lifespan distribution curve)
get_density <- function(dist_name, fit, t) {
  pars <- as.list(fit$res[,"est"])
  switch(dist_name,
    weibull = dweibull(t, shape = pars$shape, scale = pars$scale),
    lnorm   = dlnorm(t, meanlog = pars$meanlog, sdlog = pars$sdlog),
    gamma   = dgamma(t, shape = pars$shape, rate = pars$rate),
    stop("Unsupported distribution")
  )
}

sim.dens <- get_density(sim.best.name, sim.best.fit, sim.t.grid)
sim.df.plot <- data.frame(time = sim.t.grid, density = sim.dens)


```

  #Counterfactual data
```{r}
# 1) Build survival data (include right-censored individuals)
cf.surv <- cf.data %>%
  arrange(ID, Year, Age) %>%
  group_by(ID) %>%
  summarise(
    # first observed age
    first_age = min(Age, na.rm = TRUE),
    # the age when Surv == 0 appears for the first time (death event)
    death_age_event = if (any(Surv == 0, na.rm = TRUE)) Age[which(Surv == 0)[1]] else NA_real_,
    # last observed age and Surv status
    last_age = last(Age),
    last_surv = last(Surv),
    .groups = "drop"
  ) %>%
  mutate(
    # event time: if death observed, (death_age + 1 - first_age)
    # otherwise right-censored at (last_age + 1 if still survived)
    time = ifelse(
      !is.na(death_age_event),
      (death_age_event + 1) - first_age,
      (last_age + ifelse(last_surv == 1, 1, 0)) - first_age
    ),
    # status = 1 for observed death, 0 for right-censored
    status = ifelse(!is.na(death_age_event), 1, 0)
  ) %>%
  filter(is.finite(time), time > 0)

# 2) Fit several parametric survival distributions
cf.fits <- list(
  weibull = flexsurvreg(Surv(time, status) ~ 1, data = cf.surv, dist = "weibull"),
  lnorm   = flexsurvreg(Surv(time, status) ~ 1, data = cf.surv, dist = "lognormal"),
  gamma   = flexsurvreg(Surv(time, status) ~ 1, data = cf.surv, dist = "gamma")
)

# 3) Select the best model by AIC
cf.aics <- sapply(cf.fits, AIC)
cf.best.name <- names(cf.aics)[which.min(cf.aics)]
cf.best.fit  <- cf.fits[[cf.best.name]]
cat("Best distribution by AIC:", cf.best.name, "\n")

# 4) Create a fine grid of time values for plotting
cf.t.grid <- seq(min(cf.surv$time), max(cf.surv$time), length.out = 400)

#get density
cf.dens <- get_density(cf.best.name, cf.best.fit, cf.t.grid)
cf.df.plot <- data.frame(time = cf.t.grid, density = cf.dens)

```
 
  #Plot the graph
```{r}

#Combine the dataset
sim.df.plot$model <- "Simulation"
cf.df.plot$model  <- "Counterfactual"
ls.plot.df <- bind_rows(sim.df.plot, cf.df.plot)

#Plot
ggplot(ls.plot.df, aes(x = time, y = density, color = model)) +
  geom_line(linewidth = 1.2, alpha = 0.5) +
  labs(
    x = "Lifespan (years)",
    y = "Proportion",
    title = "Smoothed lifespan distributions: Simulation vs Counterfactual",
    color = "Scenario"
  ) +
  scale_color_manual(values = c(
    "Simulation"     = "#E69F00",  # 黄
    "Counterfactual" = "#0072B2"   # 蓝
  )) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12)
  )

```
  
  
#Density~LRO
  #Calculate LRO
```{r}
source("./LRO fun.R")

# For simulation dataset
sim.LRO.df <- compute_LRO_realized(sim.data)
sim.LRO.df$model <- "Simulation"

# For counterfactual dataset
cf.LRO.df <- compute_LRO_realized(cf.data)
cf.LRO.df$model <- "Counterfactual"



```
  #Plot Density~LRO
```{r}
# Combine
LRO.all <- rbind(sim.LRO.df, cf.LRO.df)

library(scales)  # for alpha()
library(tidyr)

freq.df <- LRO.all %>% 
  group_by(model, LRO) %>% 
  summarise(freq = n(), .groups = "drop") %>% 
  group_by(model) %>% mutate(prop = freq / sum(freq)) %>% # convert to proportion 
  complete(LRO = full_seq(range(LRO), 1), fill = list(freq = 0, prop = 0)) %>% # fill missing LRO values 
  ungroup()
  
ggplot(freq.df, aes(x = LRO, y = prop)) +
  geom_smooth(
    data = subset(freq.df, model == "Simulation"),
    aes(color = NULL), se = FALSE, method = "gam", formula = y ~ s(x, k = 6),
    linewidth = 1.3, color = alpha("#E69F00", 0.5)  
  ) +
  geom_smooth(
    data = subset(freq.df, model == "Counterfactual"),
    aes(color = NULL), se = FALSE, method = "gam", formula = y ~ s(x, k = 6),
    linewidth = 1.3, color = alpha("#0072B2", 0.5)  
  ) +
  labs(x = "LRO (offspring surviving to age ≥ 1, log-scaled)",
       y = "Proportion",
       title = "Smoothed proportion–LRO curves: Simulation vs Counterfactual") +
  theme_minimal(base_size = 13)


```
  




##Section 3: Model diagnostics for age-based survival and fecundity functions##

load the packages
```{r}
library(mgcv)
library(ggplot2)
```

#SURVIVAL DIAGNOSTICS
```{r}
cat("=== SURVIVAL DIAGNOSTICS ===\n")

# Fit a logistic regression model (GLM) and a smooth model (GAM)
mod.surv <- glm(Surv ~ Age, family = binomial, data = sim.data)
gam.surv <- gam(Surv ~ s(Age, k = length(unique(sim.data$Age)) - 1),
                family = binomial, data = sim.data, method = "REML")

summary(mod.surv)
summary(gam.surv)

# Predict fitted values for plotting
ages_seq <- seq(min(sim.data$Age), max(sim.data$Age))
pred_glm <- predict(mod.surv, newdata = data.frame(Age = ages_seq), type = "response")
pred_gam <- predict(gam.surv, newdata = data.frame(Age = ages_seq), type = "response")

# Plot observed vs fitted survival
plot(sim.data$Age, sim.data$Surv, pch = 1, cex = 0.7,
     xlab = "Age", ylab = "Survival probability",
     main = "Survival ~ Age diagnostics", ylim = c(0,1))
lines(ages_seq, pred_glm, col = "blue", lwd = 2)
lines(ages_seq, pred_gam, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("GLM (logistic)", "GAM (smooth)"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))

#Plot fitted VS observed residuals
plot(fitted(mod.surv), residuals(mod.surv, type = "deviance"),
     xlab = "Fitted values", ylab = "Deviance residuals")
abline(h = 0, col = "red", lty = 2)


# Basic diagnostics and model comparison
AIC(mod.surv, gam.surv)
anova(mod.surv, gam.surv, test = "Chisq")


```

#FECUNDITY DIAGNOSTICS
```{r}
cat("=== FECUNDITY DIAGNOSTICS ===\n")

# Fit Poisson regression models for fecundity (count data)
mod.fec <- glm(Offspring ~ Age, family = poisson, data = sim.data)
gam.fec <- gam(Offspring ~ s(Age, k = length(unique(sim.data$Age)) - 1),
               family = poisson, data = sim.data, method = "REML")

summary(mod.fec)
summary(gam.fec)

# Predict fitted values for plotting
pred_glm_f <- predict(mod.fec, newdata = data.frame(Age = ages_seq), type = "response")
pred_gam_f <- predict(gam.fec, newdata = data.frame(Age = ages_seq), type = "response")

# Plot observed vs fitted fecundity
plot(sim.data$Age, sim.data$Offspring, pch = 1, cex = 0.7,
     xlab = "Age", ylab = "Fecundity",
     main = "Fecundity ~ Age diagnostics")
lines(ages_seq, pred_glm_f, col = "blue", lwd = 2)
lines(ages_seq, pred_gam_f, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("GLM (Poisson)", "GAM (smooth)"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))

#Plot fitted VS observed residuals
plot(fitted(mod.surv), residuals(mod.surv, type = "deviance"),
     xlab = "Fitted values", ylab = "Deviance residuals")
abline(h = 0, col = "red", lty = 2)

# Model comparison
shapiro.test(residuals(mod.surv, type = "deviance"))
AIC(mod.fec, gam.fec)
anova(mod.fec, gam.fec, test = "Chisq")
```

