---
title: "Test-Great Tit"
author: "Wenyi Liu"
date: "2025-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the packages required and set seed
```{r}
require(doBy)
require(car)
require(mgcv) 

rm(list=ls(all=TRUE))

set.seed(123)
```

# Process the dataset

```{r}
# load the dataset
great_tit <- read.csv("Great Tit.csv")

# check the head of the columns
head(great_tit)

# Fecundity: fx = noffspring / Nx
great_tit$fx <- great_tit$noffspring / great_tit$Nx

# Survival: sx =N(x+1)/Nx The probability of indivduals with age x survive to x+1
great_tit$sx <- c(great_tit$Nx[-1] / great_tit$Nx[-nrow(great_tit)], NA)

# change NA values to zero
fx_vec <- ifelse(is.na(great_tit$fx), 0, great_tit$fx)
sx_vec <- ifelse(is.na(great_tit$sx), 0, great_tit$sx)

#Check the data
print(great_tit)

```


# Set working directory
```{r}
setwd("D:/Nian/Oxford/UG/Year 4/Year 4 project/Method/Test")
## run the utility functions
source("./Standard Graphical Pars.R")

## run the ungulate IBM
source("./Ungulate Demog Funs_GT.R")
```

#Set the simulation parameters and run the simulation
```{r}
## set the simulation parameters
init.pop.size <- 5000   
n.yrs <- 5

## run the ungulate IBM
source("./Ungulate Simulate IBM_GT.R")

## print the final population size
cat("Final population size:", tail(pop.size.t, 1), "\n")

```
take a random sample of 100 observations
we need to take a random sample because new recruits are placed at the start of the data frame 
```{r}
rows.to.keep <- sample.int(nrow(sim.data), 100)
sim.data <- sim.data[rows.to.keep,]
```



###Section 1 - Construct age-based kernels and compute population growth rates

build age-based projection kernel
```{r}
IPM.age <- mk_K_age(ages = ages,
                    sx_vec = sx_vec,
                    fx_vec = fx_vec * 0.5,
                    retain_last = TRUE)

```

calculate theoretical population growth rate

```{r}
lam.theoretical <- Re(eigen(IPM.age$K)$values[1])
cat("Theoretical λ =", lam.theoretical, "\n")
```

calculate the empirical growth rate from IBM simulation
```{r}
fit.pop.growth <- lm(log(pop.size.t) ~ seq_along(pop.size.t))
lam.empirical <- exp(coef(fit.pop.growth)[2])
cat("Empirical λ from IBM =", lam.empirical, "\n")
```


## Section 2 - Compute age-based summaries and plot ##

Stable age distribution (Normalized right eigenvector of K)
```{r}
eig_age <- eigen(IPM.age$K)
w_age <- Re(eig_age$vectors[, 1])
stable.age.dist <- w_age / sum(w_age)
```

#number (proportion) ~ lifespan
Calculate theoretical lifespan distribution
```{r}
lx_model <- cumprod(c(1, sx_vec[-length(sx_vec)]))
lx_model <- lx_model / sum(lx_model)  # normalized for plotting
```

Calculate empirical lifespan distribution
```{r}
death_tab <- table(death_records$Age_at_death)
lx_empirical <- rev(cumsum(rev(as.numeric(death_tab))))  
lx_empirical <- lx_empirical / max(lx_empirical)
ages_emp <- as.numeric(names(death_tab))
```

Plot number (proportion) ~ lifespan
```{r}
plot(ages, lx_model, type = "l", lwd = 2, col = "steelblue",
     xlab = "Age", ylab = "Proportion surviving (lx)",
     main = "Survivorship curves: IPM vs IBM")
lines(ages_emp, lx_empirical, col = "tomato", lwd = 2, lty = 2)
legend("topright", legend = c("Model (IPM)", "Empirical (IBM)"),
       col = c("steelblue", "tomato"), lwd = 2, lty = c(1, 2))
```

#fecundity~age
Calculate theoretical fecundity
```{r}
fx_model <- fx_vec   
ages_model <- ages   
```

Calculate empirical fecundity
```{r}
fecund_empirical <- aggregate(Offspring ~ Age, data = sim.data, FUN = mean)
ages_emp <- fecund_empirical$Age
fx_emp <- fecund_empirical$Offspring
```

Plot:fecundity~age
```{r}
plot(ages_model, fx_model, type = "l", lwd = 2, col = "firebrick",
     xlab = "Age", ylab = "Mean fecundity (Fx)",
     main = "Fecundity curves: IPM vs IBM",
     ylim = c(0, max(c(fx_model, fx_emp), na.rm = TRUE) * 1.1))
lines(ages_emp, fx_emp, col = "darkorange", lwd = 2, lty = 2)
legend("topright", legend = c("Model (IPM)", "Empirical (IBM)"),
       col = c("firebrick", "darkorange"), lwd = 2, lty = c(1, 2))
```



##Section 3: Model diagnostics for age-based survival and fecundity functions##

load the packages
```{r}
library(mgcv)
library(ggplot2)
```

#SURVIVAL DIAGNOSTICS
```{r}
cat("=== SURVIVAL DIAGNOSTICS ===\n")

# Fit a logistic regression model (GLM) and a smooth model (GAM)
mod.surv <- glm(Surv ~ Age, family = binomial, data = sim.data)
gam.surv <- gam(Surv ~ s(Age, k = length(unique(sim.data$Age)) - 1),
                family = binomial, data = sim.data, method = "REML")

summary(mod.surv)
summary(gam.surv)

# Predict fitted values for plotting
ages_seq <- seq(min(sim.data$Age), max(sim.data$Age))
pred_glm <- predict(mod.surv, newdata = data.frame(Age = ages_seq), type = "response")
pred_gam <- predict(gam.surv, newdata = data.frame(Age = ages_seq), type = "response")

# Plot observed vs fitted survival
plot(sim.data$Age, sim.data$Surv, pch = 1, cex = 0.7,
     xlab = "Age", ylab = "Survival probability",
     main = "Survival ~ Age diagnostics", ylim = c(0,1))
lines(ages_seq, pred_glm, col = "blue", lwd = 2)
lines(ages_seq, pred_gam, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("GLM (logistic)", "GAM (smooth)"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))

#Plot fitted VS observed residuals
plot(fitted(mod.surv), residuals(mod.surv, type = "deviance"),
     xlab = "Fitted values", ylab = "Deviance residuals")
abline(h = 0, col = "red", lty = 2)


# Basic diagnostics and model comparison
AIC(mod.surv, gam.surv)
anova(mod.surv, gam.surv, test = "Chisq")


```

#FECUNDITY DIAGNOSTICS
```{r}
cat("=== FECUNDITY DIAGNOSTICS ===\n")

# Fit Poisson regression models for fecundity (count data)
mod.fec <- glm(Offspring ~ Age, family = poisson, data = sim.data)
gam.fec <- gam(Offspring ~ s(Age, k = length(unique(sim.data$Age)) - 1),
               family = poisson, data = sim.data, method = "REML")

summary(mod.fec)
summary(gam.fec)

# Predict fitted values for plotting
pred_glm_f <- predict(mod.fec, newdata = data.frame(Age = ages_seq), type = "response")
pred_gam_f <- predict(gam.fec, newdata = data.frame(Age = ages_seq), type = "response")

# Plot observed vs fitted fecundity
plot(sim.data$Age, sim.data$Offspring, pch = 1, cex = 0.7,
     xlab = "Age", ylab = "Fecundity",
     main = "Fecundity ~ Age diagnostics")
lines(ages_seq, pred_glm_f, col = "blue", lwd = 2)
lines(ages_seq, pred_gam_f, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("GLM (Poisson)", "GAM (smooth)"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))

#Plot fitted VS observed residuals
plot(fitted(mod.surv), residuals(mod.surv, type = "deviance"),
     xlab = "Fitted values", ylab = "Deviance residuals")
abline(h = 0, col = "red", lty = 2)

# Model comparison
shapiro.test(residuals(mod.surv, type = "deviance"))
AIC(mod.fec, gam.fec)
anova(mod.fec, gam.fec, test = "Chisq")
```

